Categorical Foundations of E-Graphs With Binding

### 2025-07-10 POPL2026/paper/sections/appendix.tex
notes and popl template
comparison, semantics, enrichment fix, appendix
figures and benchmarks
fixed figures, updated contributions

\subsection{$\catname{SLat}$}
\label{sec:appendix:slat}

@@

\subsection{Pushout in $\catname{EHyp}(\Sigma)$}
\label{sec:appendix:pushout}

@@

We next explicitly construct a pushout.
Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}\}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{\\
          &\;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\\
          &\}\\
          &\cup\\
          &\{\\
          &\;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}\\
          &\}\\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
and let relations $R_{V},R_{E}$ be their transitive closures respectively.

@@

\begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = (&\\
            &\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},\\
            &\consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},\\
            &\textcolor{e-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
            &\textcolor{closed-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
        &)    
\end{align*}

@@

    This construction is analogous to how a pushout was constructed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in Fig.~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5.5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-2.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-1.5, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-0.5, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

@@

\subsection{Interpretation of $\Sigma$-terms as cospans of hypergraphs}
\label{sec:appendix:interpretation}
\begin{figure*}
  \[
  \adjustbox{width=\textwidth}{
  \tikzfig{./figures/interpretation}
  }
  \]
  \caption{Base cases for $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$}
  \label{fig:base_cases}
\end{figure*}

The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ is given in the Fig.~\ref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

\subsection{DPOI rewriting}
\label{sec:appendix:dpoi}

Consider an example of DPOI rewriting in Fig.~\ref{fig:dpoi-example} as defined in Definition~\ref{def:dpoi-e}.
The rule in the top half of the diagram is an instance of a $\beta$-reduction rule as defined for string diagrams~\cite{ghica2024stringdiagramslambdacalculifunctional}.
Note that the match satisfies all the conditions of the boundary complement as all vertices from the interface have the same parent.
First, to compute the boundary complement, we remove from the source e-hypergraph everything that has no pre-image in the interface above.
The mda-cospan for the complement then will be
\[
[u_1] \to [u_1, v_1, v_2] + [v_3] \to \mathcal{L}^{\bot} \xleftarrow{} [v_3] + [v_1,v_2] \xleftarrow{} [u_2] ~.
\]
Note how we removed strict internal interfaces of $\mathcal{L}$ from internal interfaces of $\mathcal{G}$, yielding $[u_1,v_1,v_2]$ and similarly for output interfaces.
After computing the pushout for the second half of the diagram we add strict internal interfaces of $\mathcal{R}$ but since there are none, we are done as per condition (2) of Definition~\ref{def:dpoi-e}.

\begin{figure}
\[
\adjustbox{width=\linewidth}{
\tikzfig{./figures/dpo_example}
}
\]
\caption{$\beta$-rule DPOI example.}
\label{fig:dpoi-example}
\end{figure}

>>>>

\section{$\catname{SLat}$}
\label{sec:appendix:slat}

@@

\section{Pushout in $\catname{EHyp}(\Sigma)$}
\label{sec:appendix:pushout}

@@

We next explicitly construct a pushout.
Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}\}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{
          \;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
          \}\\
          &\cup\\
          &\{
          \;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
          \}\\
          &\cup\\
          &\{
          \;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}
          \}\\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
and let relations $R_{V},R_{E}$ be their transitive closures respectively.

@@

\begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = &(
            \sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},\\
            &\consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},\\
            &\textcolor{e-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
            &\textcolor{closed-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}})    
\end{align*}

@@

    This construction is analogous to how a pushout was constructed in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in Fig.~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
    \adjustbox{scale=0.8}{
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-4, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-3, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-2.5, 3.5) {};
	\end{pgfonlayer}
    }
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

@@

\section{Interpretation of $\Sigma$-terms as cospans of hypergraphs}
\label{sec:appendix:interpretation}
\begin{figure*}
  \[
  \adjustbox{width=\textwidth}{
  \tikzfig{./figures/interpretation}
  }
  \]
  \caption{Base cases for $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$}
  \label{fig:base_cases}
\end{figure*}

The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ are given in \autoref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

\section{DPOI rewriting}
\label{sec:appendix:dpoi}

Consider an example of DPOI rewriting in \autoref{fig:dpoi-example} as defined in Definition~\ref{def:dpoi-e}.
The rule in the top half of the diagram is an instance of a $\beta$-reduction rule as defined for string diagrams~\cite{ghica2024stringdiagramslambdacalculifunctional}.
Note that the match satisfies all the conditions of the boundary complement as all vertices from the interface have the same parent.
First, to compute the boundary complement, we remove from the source e-hypergraph everything that has no pre-image in the interface above.
The mda-cospan for the complement then will be
\[
[u_1] \to [u_1, v_1, v_2] + [v_3] \to \mathcal{L}^{\bot} \xleftarrow{} [v_3] + [v_1,v_2] \xleftarrow{} [u_2] ~.
\]
Note how we removed strict internal interfaces of $\mathcal{L}$ from internal interfaces of $\mathcal{G}$, yielding $[u_1,v_1,v_2]$ and similarly for output interfaces.
After computing the pushout for the second half of the diagram we add strict internal interfaces of $\mathcal{R}$ but since there are none, we are done as per condition (2) of \autoref{def:dpoi-e}.

\begin{figure}
    \ifdefined\ONECOLUMN
    \[
    \adjustbox{width=0.6\linewidth}{
    \tikzfig{./figures/dpo_example}
    }
    \]
    \else
    \[
\adjustbox{width=\linewidth}{
\tikzfig{./figures/dpo_example}
}
\]
    \fi
    \caption{$\beta$-rule DPOI example.}
    \label{fig:dpoi-example}
\end{figure}

### 2025-07-10 POPL2026/paper/sections/combinatorial_semantics.tex
fixed figures, updated contributions

\begin{definition}
	We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{e-color}{<}, \textcolor{closed-color}{<},\consistency)$ where
	\begin{itemize}
		\item $V$ is a set of vertices.
		\item $E = \colorbox{yellow}{E} \cup \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$ is a set of hyperedges that is formed of three disjoint sets of hyperedges defined below.
		\item $s,t : E \to V^{*}$ are source and target functions.
		\item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
		\item $\textcolor{e-color}{<} \subseteq \textcolor{e-color}{E} \times (V + E)$ and $\textcolor{closed-color}{<} \subseteq \textcolor{closed-color}{E} \times (V + E)$ with $\textcolor{e-color}{E}, \textcolor{closed-color}{E} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{e-color}{<^{\mu}}$ and $\textcolor{closed-color}{<^{\mu}}$.
		      For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{e-color}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{closed-color}{<^{\mu}} y_{i+1}\; \}$.
		      We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
		      Each of these relations should satisfy the following
		      \begin{enumerate}
			      \item each set of predecessors consists exclusively of hierarchical edges;
			      \item each $x$ has at most one immediate predecessor;
			      \item edges $e$ such that there is no $x$ such that $e \textcolor{e-color}{<^{\mu}} x$ or $e \textcolor{closed-color}{<^{\mu}} x$ are labelled;
			      \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{e-color}{<^{\mu}} e$ iff $e' \textcolor{e-color}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{closed-color}{<^{\mu}}$.
		      \end{enumerate}
		\item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{e-color}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{e-color}{<^{\mu}}(v)$ and $p \textcolor{e-color}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
		      We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{e-color}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{e-color}{<^{\mu}} e\}$.
	\end{itemize}
	$\textcolor{e-color}{E}$ and $\textcolor{closed-color}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$.
	When writing $x \textcolor{closed-color}{<} y$ (or $x \textcolor{e-color}{<} y$) we may refer to $x$ as a predecessor of $y$ and refer to $y$ as a successor of $x$.
	% we do not use this
	% Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{e-color}{<}, \textcolor{closed-color}{<}$ and~$\consistency$.
\end{definition}

>>>>

\begin{definition}
	We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{e-color}{<}, \textcolor{closed-color}{<},\consistency)$ where
	\begin{itemize}
		\item $V$ is a set of vertices.
		\item $E = \colorbox{yellow}{E} \cup \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$ is a set of hyperedges that is formed of three disjoint sets of hyperedges defined below.
		\item $s,t : E \to V^{*}$ are source and target functions.
		\item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
		\item $\textcolor{e-color}{<} \subseteq \textcolor{e-color}{E} \times (V + E)$ and $\textcolor{closed-color}{<} \subseteq \textcolor{closed-color}{E} \times (V + E)$ with $\textcolor{e-color}{E}, \textcolor{closed-color}{E} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{e-color}{<^{\mu}}$ and $\textcolor{closed-color}{<^{\mu}}$.
		      For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{e-color}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{closed-color}{<^{\mu}} y_{i+1}\; \}$.
		      We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
		      Each of these relations should satisfy the following
		      \begin{enumerate}
			      \item each set of predecessors consists exclusively of hierarchical edges;
			      \item each $x$ has at most one immediate predecessor;
			      \item edges $e$ such that there is no $x$ such that $e \textcolor{e-color}{<^{\mu}} x$ or $e \textcolor{closed-color}{<^{\mu}} x$ are labelled;
			      \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{e-color}{<^{\mu}} e$ iff $e' \textcolor{e-color}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{closed-color}{<^{\mu}}$.
		      \end{enumerate}
		\item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{e-color}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{e-color}{<^{\mu}}(v)$ and $p \textcolor{e-color}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
		      We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{e-color}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{e-color}{<^{\mu}} e\}$.
	\end{itemize}
	$\textcolor{e-color}{E}$ and $\textcolor{closed-color}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$.
	When writing $x \textcolor{closed-color}{<} y$ (or $x \textcolor{e-color}{<} y$) we may refer to $x$ as the predecessor of $y$ and refer to $y$ as a successor of $x$.
	% we do not use this
	% Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{e-color}{<}, \textcolor{closed-color}{<}$ and~$\consistency$.
\end{definition}

### 2025-07-10 POPL2026/paper/sections/introduction.tex
fixed figures, updated contributions

\citet{ghica2024equivalencehypergraphsegraphsmonoidal} shows that e-graphs over signature $\Sigma$ can be represented as morphisms of a free cartesian symmetric monoidal semilattice-enriched category over the same signature.
We argue that restricting the categorical domain to free \textbf{closed} symmetric monoidal semilattice-enriched categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding (using the closed structure) and equivalence (using the semilattice enrichment) between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of them, \emph{closed e-hypergraphs}.
We equip closed e-hypergraphs with a suitable notion of DPO rewriting, and show that this notion of rewriting is sound and complete with respect to rewriting of morphism terms.

>>>>

\citet{ghica2024equivalencehypergraphsegraphsmonoidal} shows that e-graphs over signature $\Sigma$ can be represented as morphisms of a free cartesian symmetric monoidal semilattice-enriched category over the same signature.
We argue that restricting the categorical domain to free \textbf{closed} symmetric monoidal semilattice-enriched categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding (using the closed structure) and equivalence (using the semilattice enrichment) between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of them, \emph{closed e-hypergraphs}.
First, we define closed enriched $\Sigma$-terms which generate a theory with equivalences and bindings.
Then we define what e-hypergraphs are and provide the interpretation of closed enriched $\Sigma$-terms as cospans of e-hypergraphs (and by doing this we also formalise the corresponding string diagrams as concrete combinatorial objects).
Next we formulate the DPO rewriting for e-hypergraphs and show that this notion of rewriting is sound and complete with respect to rewriting of closed enriched $\Sigma$-terms modulo SMC equations.
We finish the paper with a description of what benefits the e-hypergraph representation may bring for equality saturation in the context of $\lambda$-calculus with explicit substitution compared to conventional (slotted) e-graphs.

### 2025-07-10 POPL2026/paper/sections/soundness.tex
mass format
fixed figures, updated contributions

\begin{definition}
\label{def:rewrite}
    We say that a $\Sigma$-term $f$ rewrites to a $\Sigma$-term $g$ modulo SMC laws via a rewrite rule $\langle l, r \rangle$ (notation, $\leadsto_{\langle l, r \rangle}$) if they are representable as
    $
    f = c_{1};(id_{k} \otimes l);c_{2} $ and $ g = c_{1};(id_{k} \otimes r);c_{2}
    $.
\end{definition}
This notion generalises to rewriting of closed $\Sigma$-terms
\begin{definition}
    For closed-$\Sigma$-terms $f,g$, $f \leadsto_{\langle l, r \rangle} g$ modulo SMC laws if they are representable as either in Definition~\ref{def:rewrite}, or as 
    $
    f = c_{1};(id_{k} \otimes d);c_{2} $ and $ g = c_{1};(id_{k} \otimes e);c_{2}
    $
    and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}
\begin{lemma}
\label{lemma:normal_form}
    Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
    $
    f_{1} + \ldots + f_{n}
    $ modulo SMC laws and $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} where none of $f_{i}$ contain the join operator.
\end{lemma}
% \begin{proof}
% Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
% Most notably, we can also escape lambda abstractions by using distributivity law~\ref{law:distributivity}.
% \end{proof}
This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
    For closed $\Sigma^{+}$-term $f,g$, $f \leadsto_{\langle l, r \rangle}$ SMC laws, $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} if they are representable as
    $
    f = f_{1} + \ldots + f_{i} + \ldots + f_{n} $ and $ g = f_{1} + \ldots + f_{j} + \ldots + f_{n}
    $
    such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}
Next, to formulate the correspondence between rewriting systems we first define interpretation $\llbracket - \rrbracket$ of closed $\Sigma^{+}$ terms in $\WellTypedMdaEcospans$.
It follows by first defining the interpretation of operators by extending the interpretation of plain $\Sigma$ terms as morphisms in $\MdaCospans$ as given in Appendix~\ref{sec:appendix:interpretation}.
The only missing cases, $\llbracket \textsf{ev}_{A,B} \rrbracket$ and $\llbracket \Lambda_{A,B,C} \rrbracket$, are given in Fig.~\ref{fig:ev_and_lambda}.

@@

>>>>

\begin{definition}[(Closed) $\Sigma$-term rewrite rule]
A $\Sigma$-term rewrite rules is a pair $\langle l, r \rangle$ where $l$ and $r$ are $\Sigma$ terms with matching inputs and outputs (as lists of input and output types).
When $l$ and $r$ are closed $\Sigma$-terms, we call such a rewrite rule closed.
\end{definition}

\begin{definition}
	\label{def:rewrite}
	We say that a $\Sigma$-term $f$ rewrites to a $\Sigma$-term $g$ modulo SMC laws via a rewrite rule $\langle l, r \rangle$ (notation, $\leadsto_{\langle l, r \rangle}$) if they are representable as
	$
		f = c_{1};(id_{k} \otimes l);c_{2} $ and $ g = c_{1};(id_{k} \otimes r);c_{2}
	$.
\end{definition}
This notion generalises to rewriting of closed $\Sigma$-terms.

\begin{definition}
	For closed-$\Sigma$-terms $f,g$, and a closed rewrite rule $\langle l, r \rangle$, $f \leadsto_{\langle l, r \rangle} g$ modulo SMC laws if they are representable as either in Definition~\ref{def:rewrite}, or as
	$
		f = c_{1};(id_{k} \otimes d);c_{2} $ and $ g = c_{1};(id_{k} \otimes e);c_{2}
	$
	and $d \leadsto_{\langle l, r \rangle} e$
\end{definition}
\begin{lemma}
	\label{lemma:normal_form}
	Every closed $\Sigma^{+}$ term $f$ can be equivalently represented as
	$
		f_{1} + \ldots + f_{n}
	$ modulo SMC laws and $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} where none of $f_{i}$ contain the join operator.
\end{lemma}
% \begin{proof}
% Such forms are given by orienting $\catname{SLat}$-equations such that $+$ gets propagated to the top, as, for example $f \otimes (g + h) = f \otimes g + g \otimes h$.
% Most notably, we can also escape lambda abstractions by using distributivity law~\ref{law:distributivity}.
% \end{proof}
This takes us to the following definition of rewriting for closed $\Sigma^{+}$-terms.
\begin{definition}
	For closed $\Sigma^{+}$-term $f,g$, and a closed $\Sigma$-term rewrite rule $\langle l, r \rangle$ $f \leadsto_{\langle l, r \rangle}$ modulo SMC laws, $\catname{SLat}$-equations and distributivity law~\ref{law:distributivity} if they are representable as
	$
		f = f_{1} + \ldots + f_{i} + \ldots + f_{n} $ and $ g = f_{1} + \ldots + f_{j} + \ldots + f_{n}
	$
	such that there is a permutation $\sigma$ such that there exist indices $i,j$ and $\sigma(f_{i}) \leadsto f_{j}$.
\end{definition}
Note that the rewrite rule consist of closed $\Sigma$-terms and not closed $\Sigma^{+}$-terms, i.e. neither $l$ nor $r$ contain any $+$.
This is intentional as there rewrite rules are supposed to be domain-specific (like a rule $\langle 2 \otimes id;times, id \otimes 1; <\;\;< \rangle$) and not related to the free enrichment.

Next, to formulate the correspondence between rewriting systems we first define interpretation $\llbracket - \rrbracket$ of closed $\Sigma^{+}$ terms in $\WellTypedMdaEcospans$.
It follows by first defining the interpretation of operators by extending the interpretation of plain $\Sigma$ terms as morphisms in $\MdaCospans$ as given in Appendix~\ref{sec:appendix:interpretation}.
The only missing cases, $\llbracket \textsf{ev}_{A,B} \rrbracket$ and $\llbracket \Lambda_{A,B,C} \rrbracket$, are given in Fig.~\ref{fig:ev_and_lambda}.

@@

Practically, this quotienting means that we perform DPO rewrites modulo the equations of SMC and modulo the law~\ref{law:distributivity}.
We can either rewrite the normal forms as mentioned above that contain all possible redexes, or we can genuinely search for redexes modulo these laws by traversing the e-hypergraph through both types of boxes when searching for a redex, which makes this representation practically viable even though it does not absorb all the possible equations.
